`flippant` -- An `R` package for the automated analysis of dithionite scramblase

### 2015-09-24 inst/manuscript/manuscript.rmd
Integrate edits by Birgit Ploier;

# Introduction
Scramblases are proteins mediating the mass-action driven equilibration of 
lipids between the leaflets of lipid bilayers that constitute biological membranes
[@waechter_flippases_2013]. Independent of energy equivalents such as ATP they 
facilitate this ``flipping'' by providing a mechanism of transition through the 
hydrophobic environment of membranes to polar lipid head groups, thus reducing 
the massive energetic cost that process would carry otherwise 
[@kornberg_inside-outside_1971]. The underlying molecular mechanisms remain 
largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified **(I was looking for a recent 
citation that clearly spelled that out - in vain. Do you have a proposal?)**. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, we speculate that lipid scrambling may, in fact, 
be an intrinsic property of G-protein coupled receptors in general 
[@goren_constitutive_2014].

@@

##Scramblase Assay Analysis
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart 
of the analysis of dithionite scramblase assays, allowing for the visual 
comparison of specific scramblase activity between interrogated experimental 
series, which may e.g. be representative of protein preparations including 
mutant protein forms or crude extracts with or without depletion of a candidate
scramblase. 
```{r, code = formatR::usage('scramblaseAssayPlot'), eval = FALSE}
```
allows for the complete generation of the PPR plot in publication quality from 
the user assembled spreadsheet of input data (`x`), including automatically 
added graphical representations of $\tau$, the PPR value equal to the fit 
constant $\alpha$, which implies all liposomes on average carrying one 
scramblase and 63% have one or more (i.e. are active). The behavior of the 
function may be adjusted to use scramblase-mediated fluorescence reduction 
normalized to either the maximum measurement in the series (`scaleTo = "data"`)
or the plateau of an exponential fit to it (`scaleTo = "model"`; for reasoning
see above) and whether or not to require the exponential curve fitted to the 
data to go through the origin.

Numeric scramblase assay results such as $\tau$/$\alpha$ may be generated in 
tabular form using
```{r, code = formatR::usage('scramblaseAssayStats'), eval = FALSE}
```
with the same modifying arguments to the function as for `scramblaseAssayPlot`.

@@

Scramblase activity analysis as described above is performed on the data set with
the following calls.
```{r, eval = FALSE}
# Plot PPR plot (Figure 1B)
scramblaseAssayPlot("inputTable.txt") + xlim(0, 2)
# Write out results in tabular manner
scramblaseAssayStats("inputTable.txt")
```
```{r, echo = FALSE}
setwd(tmpDir)
max_PPR <- read.delim(
    "inputTable.txt",
    check.names = FALSE) %>%
  flippant:::calculatePpr() %>%
  max(
    na.rm = TRUE) %>%
  round(
    digits = 2
  )
```
The first call results in the protein to phospholipid ratio (PPR) plot shown in
Figure 1B. Using the customization facilities provided by `ggplot2` 
[@wickham_ggplot2_2009] as a plotting system the plot's x-axis is trimmed to
avoid visual compression of all data by a single data point from the wild type 
rhodopsin series in experiment 1, where $\text{PPR}=`r max_PPR`$. Dashed lines
are added automatically as graphical representations of the fit constant 
$\alpha$ or $\tau$, but are difficult to evaluate visually.

Recapitulation of the calculations underlying PPR plot generation by 
`scramblaseAssayPlot` in a call to `scramblaseAssayStats` produces tabular 
output represented in Table 1, which includes series-specific fit constants 
$\alpha$ in numeric and easily accessible form.
```{r, echo = FALSE}
setwd(tmpDir)
ppr_results <- scramblaseAssayStats("inputTable.txt")
ppr_results_tau_average <- ppr_results %>%
  split(ppr_results$`Experimental Series`) %>% 
  sapply(function(y){mean(y$`Fit Constant`, na.rm = TRUE)}) %>%
  round(2)
ppr_results_tau_average_names <- names(ppr_results_tau_average) %>%
  head(n = length(ppr_results_tau_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(names(ppr_results_tau_average), n = 1))
ppr_results_tau_average_print <- ppr_results_tau_average %>%
  unname() %>%
  head(n = length(ppr_results_tau_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(ppr_results_tau_average, n = 1))
knitr::kable(
  ppr_results,
  row.names = FALSE,
  caption = "**Tabular output of `flippant`-based reanalysis of data in Figure 4C/D from @ploier_perturbed_submitted.** An `ASCII`-version of this output is produced by calling `scramblaseAssayStats`.")
```

Mean $\tau$ values (or fit constants $\alpha$) correspond with `r ppr_results_tau_average_print` for `r ppr_results_tau_average_names`, 
respectively, very well to the prior manual analysis by 
@ploier_perturbed_submitted, where $\tau \approx 0.12$ was reported for mutant 
and $\tau \approx 0.26$ for wild type rhodopsin. Based on the manual analysis of
the data we conclude in @ploier_perturbed_submitted that the mutant rhodopsins
investigated are reconstituted into proteoliposomes in monomeric form, contrary 
to the wild type, which is reconstituting as a homodimer. `flippant` analysis of
the data, shows identical trends and close matches of the results, supporting 
identical conclusions.

>>>>

# Introduction
Scramblases are proteins mediating the bidirectional, mass-action driven 
equilibration of lipids between the leaflets of lipid bilayers that constitute 
biological membranes [@waechter_flippases_2013]. Independent of energy 
equivalents such as ATP they facilitate this ``flipping'' by providing a 
mechanism of transition through the hydrophobic environment of membranes to 
polar lipid head groups, thus reducing the massive energetic cost that process 
would carry otherwise [@kornberg_inside-outside_1971]. The underlying molecular 
mechanisms remain largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified **(I was looking for a recent 
citation that clearly spelled that out - in vain. Do you have a proposal?)**. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, we speculate that lipid scrambling may, in fact, 
be an intrinsic property of G-protein coupled receptors in general 
[@goren_constitutive_2014].

@@

##Scramblase Assay Analysis
Plotting protein to phospholipid ratio (PPR) against the likelihood for a given
proteoliposome to contain one or more scramblases, the PPR plot is at the heart 
of the analysis of dithionite scramblase assays, allowing for the visual 
comparison of specific scramblase activity between interrogated experimental 
series, which may e.g. be representative of protein preparations including 
mutant protein forms or crude extracts with or without depletion of a candidate
scramblase. 
```{r, code = formatR::usage('scramblaseAssayPlot'), eval = FALSE}
```
allows for the complete generation of the PPR plot in publication quality from 
the user assembled spreadsheet of input data (`x`), including automatically 
added graphical representations the PPR value equal to the fit 
constant $\alpha$, which implies all liposomes on average carrying one 
scramblase and 63% have one or more (i.e. are active). The behavior of the 
function may be adjusted to use scramblase-mediated fluorescence reduction 
normalized to either the maximum measurement in the series (`scaleTo = "data"`)
or the plateau of an exponential fit to it (`scaleTo = "model"`; for reasoning
see above) and whether or not to require the exponential curve fitted to the 
data to go through the origin.

Numeric scramblase assay results such as $\alpha$ may be generated in 
tabular form using
```{r, code = formatR::usage('scramblaseAssayStats'), eval = FALSE}
```
with the same modifying arguments to the function as for `scramblaseAssayPlot`.

@@

Scramblase activity analysis as described above is performed on the data set with
the following calls.
```{r, eval = FALSE}
# Plot PPR plot (Figure 1B)
scramblaseAssayPlot("inputTable.txt") + xlim(0, 2)
# Write out results in tabular manner
scramblaseAssayStats("inputTable.txt")
```
```{r, echo = FALSE}
setwd(tmpDir)
max_PPR <- read.delim(
    "inputTable.txt",
    check.names = FALSE) %>%
  flippant:::calculatePpr() %>%
  max(
    na.rm = TRUE) %>%
  round(
    digits = 2
  )
```
The first call results in the protein to phospholipid ratio (PPR) plot shown in
Figure 1B. Using the customization facilities provided by `ggplot2` 
[@wickham_ggplot2_2009] as a plotting system the plot's x-axis is trimmed to
avoid visual compression of all data by a single data point from the wild type 
rhodopsin series in experiment 1, where $\text{PPR}=`r max_PPR`$. Dashed lines
are added automatically as graphical representations of the fit constant 
$\alpha$, but are difficult to evaluate visually.

Recapitulation of the calculations underlying PPR plot generation by 
`scramblaseAssayPlot` in a call to `scramblaseAssayStats` produces tabular 
output represented in Table 1, which includes series-specific fit constants 
$\alpha$ in numeric and easily accessible form.
```{r, echo = FALSE}
setwd(tmpDir)
ppr_results <- scramblaseAssayStats("inputTable.txt")
ppr_results_alpha_average <- ppr_results %>%
  split(ppr_results$`Experimental Series`) %>% 
  sapply(function(y){mean(y$`Fit Constant`, na.rm = TRUE)}) %>%
  round(2)
ppr_results_alpha_average_names <- names(ppr_results_alpha_average) %>%
  head(n = length(ppr_results_alpha_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(names(ppr_results_alpha_average), n = 1))
ppr_results_alpha_average_print <- ppr_results_alpha_average %>%
  unname() %>%
  head(n = length(ppr_results_alpha_average) - 1) %>%
  paste(collapse = ", ") %>%
  paste("and") %>%
  paste(tail(ppr_results_alpha_average, n = 1))
knitr::kable(
  ppr_results,
  row.names = FALSE,
  caption = "**Tabular output of `flippant`-based reanalysis of data in Figure 4C/D from @ploier_perturbed_submitted.** An `ASCII`-version of this output is produced by calling `scramblaseAssayStats`.")
```

Mean values of fit constant $\alpha$) correspond with 
`r ppr_results_alpha_average_print` for `r ppr_results_alpha_average_names`, 
respectively, very well to the prior manual analysis by 
@ploier_perturbed_submitted, where $\alpha \approx 0.13$ was reported for mutant 
and $\alpha \approx 0.26$ for wild type rhodopsin. Based on the manual analysis 
of the data we conclude in @ploier_perturbed_submitted that the mutant 
rhodopsins investigated are reconstituted into proteoliposomes in monomeric 
form, contrary to the wild type, which is reconstituting as a homodimer. 
`flippant` analysis of the data, shows identical trends and close matches of the
results, supporting identical conclusions.
